An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-10-03 paper.tex
Complete §7.1 (successful author)

@@

% ICWSM 2022 paper section 4.3
The statistical analysis shows that influence of draft authors in the email
networks does impact the possibility of a draft getting adopted
(\S\ref{subsec:results}, Table \ref{tbl:res} \& \ref{tbl:resstat}). This
might hint at the ability of participants who hold a domain expertise to be
able to engage better with the community. Several WG chairs are already in
the top percentile influential category in both the email and co-authorship
networks before taking up these leadership roles, which further elevates
after taking up such leadership roles (Figure
\ref{fig:wg_chair_before_after}).

Most importantly the statistical analysis shows that being affiliated with
a prominent organisation positively impacts the chances of a draft in
getting adopted by a WG, thereby directly driving the innovation process
(Table \ref{tbl:res} \& \ref{tbl:resstat}).

>>>>

\todo{introductory remarks}

@@

We have identified that influential authors write more drafts
(\S\ref{subsec:impact_influence}), but are these drafts more likely to
see success?  That is, does the presence of a well-connected, influential,
author make it more likely that an Internet-draft is adopted by a working
group for review and development\footnote{There are occasional document,
known as Area Director sponsored drafts, that are published as standards
track RFCs without going through the working group process, but these too
rare ($\sim$3\% of RFCs) to significantly affect our analysis.} and
eventually published as a standards track RFC.  Using information from the
IETF Datatracker we compile a dataset for the \emph{adoption} of a draft by
a working group, comprising 11632 drafts, and employ logistic regression to
analyse factors that affect working group adoption of draft.

%..................................................................................................
\pb{Methodology:}
We compute a set of features that may impact adoption or publication for
each draft, most of them motivated by the characteristics of influential
participants, including the following feature groups: 
 (1) \emph{text}, we derive a term frequency--inverse document
     frequency (TF-IDF) \cite{salton1983introduction} weighted bag of words vector
     representation of all email text of messages that either explicitly
     mention a draft or are part of a thread where the subject line
     explicitly mentions a draft; 
 (2) \emph{communication patterns}, i.e., volumes of incoming/outgoing communication
     between draft authors and IETF participants of varying experience levels;
 (3) Betweenness \emph{centrality} for authors;
 (4) \emph{number of emails} sent by authors;
 (5) \emph{number of drafts} submitted by authors;
 (6) \emph{length of participation} duration of authors;
 (7) \emph{number of authors};
 (8) \emph{number of IETF areas} in which the author is active;
 (9) \emph{proportion of top authors} in the top 10 and top 20 percent of
     influential participants;
(10) \emph{topic entropy ($\eta$)} scores for authors;
(11) \emph{author affiliations} for each of the 20 most influential
     organizations (20 boolean features);
(12) \emph{number of mailing lists} that each author participates in,
(13) \emph{number of years active}.

For numerical features that relate to each author individually instead of
the draft as a whole (e.g., \emph{number of emails sent}, \emph{centrality},
or \emph{number of years active}, the feature defined for each author
separately) in three variants: (i) for the least influential author, (ii)
for the most influential author, and (iii) as the average value across all
authors of the draft. We compute individual author features using
information from the 5-year period prior to draft submission.

Regarding the text based features, we found that the models tend to assign
high weights to words such as surnames of active contributors, names of
prolific working groups, and common technical terms. While this does make
the models perform slightly better, such terms should not be relevant for
solving the task as they do not model the relevant part of the conversation.
To explore how models would behave in a scenario wherein such words were
not available, we construct a domain-specific stop word list, consisting
of: (i) one all last names from the Datatracker, (ii) all working group
names from the Datatracker, and (iii) technological jargon terms obtained
from the web\footnote{\url{https://www.computerhope.com/jargon.htm}, we
used a union of the \emph{Internet terms}, \emph{network terms}, and
\emph{security terms} categories.}. We remove from this list any terms
appearing in top 5,000 English terms (e.g., sometimes people's names can have
identical surface forms as some of the terms relevant to organisational
activities). We will refer to this variant as \emph{text (S)}.

To gain empirical insights into which features provide better prediction
results, models are run using a feature group alone variant and a variant
combining each feature group with the \emph{text} features. This approach
was motivated by the finding that \emph{text}, even though conceptually
quite different from the rest of the feature set and not our main focus,
does provide surprisingly strong results on its own. Thus, we wanted to
empirically investigate in more detail how well it complements the rest of
the graph-based features. Moreover, we compare all models to a baseline
which simply assigns all examples to the majority class.

We split the data into \emph{training} (70\%), \emph{development} (10\%),
and \emph{test} (20\%) subsets. As a scoring function, we use the F1 score
macro averaged across the positive and negative classes (but we also report
area under the curve (AUC), precision, and recall). We train the models on
the training set, optimise hyper-parameters on the development set, and
report final scores on the test set. The final scores are those obtained by
the model variant that fared best on the development set. We use logistic
regression implemented in \texttt{scikit-learn} \cite{scikit-learn}, given
that it is widely used and well interpretable. While we do not perform
explicit feature selection, we do have implicit feature selection through
L1 regularisation. We consider the regularisation strength a hyper-parameter
and consider values from $[2^{-7}, 2^{-6}, ..., 2^6]$ on the development
set. As our data set is quite imbalanced (in a roughly 4:1 ratio, 17\% is
\emph{adopted}), we use different class weights during optimization to
counteract this. We employ a non-parametric random shuffling test
\cite{yeh-2000-accurate} to check statistical significance of score
differences against the baseline. Table~\ref{tbl:res} summarises prediction
results, which will be discussed in the next section.

\renewcommand{\tabcolsep}{0.5cm}
\begin{table*}
\centering
\begin{tabular}{lcccccccc}
\toprule
                        & \multicolumn{4}{c}{$\neg \mathit{TXT}$}
                        & \multicolumn{4}{c} {$\mathit{TXT}$} \\
\midrule
                       & AUC & F1 & P & R & AUC & F1 & P & R \\
\midrule
Baseline               &  .500 & .445 & .401 & .500  &  .500 & .445 & .401 & .500 \\
\midrule
Comm. patterns         &  .644 & .583 & .566 & .601 &  .723 & .626 & .609 & .644 \\
Centrality             &  .655 & .577 & .573 & .581 &  .717 & .639 & .625 & .653 \\
Email count            &  .632 & .566 & .555 & .577 &  .718 & .650 & .639 & .662 \\
N. years active        &  .608 & .569 & .555 & .584 &  .702 & .610 & .594 & .626 \\
Number of authors      &  .578 & .536 & .529 & .544 &  .697 & .631 & .621 & .641 \\
Proportion in top      &  .663 & .594 & .580 & .609 &  .728 & .634 & .614 & .655 \\
Draft count            &  .500 & .445 & .401 & .500 &  .696 & .630 & .624 & .636 \\
N. Areas               &  .634 & .572 & .557 & .588 &  .721 & .639 & .622 & .657 \\
N. Mailing lists       &  .634 & .570 & .557 & .584 &  .718 & .641 & .630 & .653 \\
Affiliations           &  .605 & .584 & .576 & .592 &  .708 & .626 & .609 & .643 \\
Topic entropy ($\eta$) &  .622 & .563 & .550 & .577 &  .715 & .642 & .628 & .657 \\
Text                   & - & - & -  & -  &  .681 & .624 & .622 & .627 \\
Text (S)               & - & - & -  & -  &  .667 & .598 & .593 & .614 \\
\midrule
All feats              &  .692 & .624 & .600 & .651 &  .744 & .645 & .625 & .666 \\
All feats (S)          &  .692 & .624 & .600 & .651 &  .726 &  .636 &  .627 &  .664 \\
\bottomrule
\end{tabular}
\caption{
  Results for predicting adoption. Each row presents the scores of a model
  using the corresponding feature group either alone ($\neg \mathit{TXT}$)
  or combined with the Text features ($\mathit{TXT}$). \emph{All feats}
  denotes all features except the text. Rows labeled with (S)  use the
  \emph{Text (S)} variant of the text features.
}
\label{tbl:res}
\end{table*}

Next, we wanted to confirm these results at the level of individual
features. To this end, we inspect the statistically significant
coefficients learned by the model corresponding to each individual feature.
For this experiment, we use a slightly different setup. We do not consider
the \emph{text} features\footnote{There are thousands of text features
(terms), making it prohibitively complex to include them. Moreover, they
are not central to our research questions about the IETF community.}. We
begin by first applying the Variance Inflation Factor (VIF) to exclude all
features with $\mathit{VIF} > 5$. This, to an extent, mitigates the
collinearity that we know exists as some features are by construction
highly correlated. We then standardise each remaining feature and then fit
a Logistic Regression model from the \texttt{statsmodels} package
\cite{seabold2010statsmodels} on the entire data set.
Table~\ref{tbl:resstat} presents the results of this experiment.

\renewcommand{\tabcolsep}{0.3cm}
\begin{table}
  \centering
  \begin{tabular}{lccc}
    \toprule
      Feature       & Weight  & P-value \\
    \midrule
      Centrality of most influential  & 0.1691 &   0.000 \\
      Centrality of least influential & -0.0698 &  0.048 \\
      Proportion of authors in top 10  & 0.1520 &  0.000 \\
    \midrule
      Has a Cisco author  & 0.1301 &  0.000 \\
      Has an AT\&T author  & 0.1003 &  0.000 \\
      Has a Juniper author  & 0.0738 &  0.000 \\
      Has a China Mobile author & -0.0697  & 0.000 \\
      Has an Alcatel-Lucent author  & 0.0474 &  0.022 \\
    \midrule
      Topic entropy max & 0.1728 & 0.000 \\
      Topic entropy min & -0.1135 &  0.002 \\
    \bottomrule
  \end{tabular}
  \caption{
    Weights of the model that are statistically significant at $p \leq 0.05$
    grouped by similarity of features -- author influence measures vs.\
    affiliation features vs.\ topic entropy ($\eta$).
  }
  \label{tbl:resstat}
\end{table}

%..................................................................................................
\pb{Predicting Adoption:}
To confirm empirical implications from Section \ref{subsec:impact_influence},
and provide a more in-depth look at the features, we perform statistical
analysis summarised in Table~\ref{tbl:resstat}. The most interesting
insight is that the coefficients corresponding to the \emph{Centrality of
the most influential author} and \emph{Proportion of authors in the top 10
percentile in influence}, are, indeed, positive and among the largest in
absolute value. This highlights that among the people who are influential
in the email networks, there is a substantial number of individuals who are
proficient in contributing to successful drafts.

Another interesting observation is that author affiliations (which are
also indirectly connected to influential participants, as demonstrated in 
\S\ref{subsec:impact_influence}, Figure \ref{fig:coauthor_affiliation_network2019_otheryears})
are an important feature group, particularly affiliations such as \textit{Cisco}
and \textit{AT\&T}. Moreover, topic entropy is also an important feature
group, which was also shown to be related to influential participants in
our analysis from \S\ref{subsec:impact_influence}, Figure \ref{fig:topic_diversity}.

%..................................................................................................
% ICWSM 2022 paper section 4.3
\pb{Summary:}
The statistical analysis shows that influence of draft authors in the email
networks does impact the possibility of a draft getting adopted (Tables
\ref{tbl:res} \& \ref{tbl:resstat}). This might hint at the ability of
participants who hold a domain expertise to be able to engage better with
the community. Several WG chairs are already in the top percentile
influential category in both the email and co-authorship networks before
taking up these leadership roles, which further elevates after taking up
such leadership roles (Figure \ref{fig:wg_chair_before_after}).  Our
analysis also shows that being affiliated with a prominent organisation
positively impacts the chances of a draft in getting adopted by a working
group, thereby directly driving the innovation process (Table \ref{tbl:res}
\& \ref{tbl:resstat}).
